---
title: "cyclist case study"
author: "harsh haldankar"
date: "2023-1-11"
output:
  html_document: default
---
 
#### Load the packages

```{r}
install.packages('tidyverse')
install.packages("skimr")
library(tidyverse)
library(dplyr)
library(skimr)
library(lubridate)
library(ggplot2)
```

#### getting the path

```{r}
getwd()
```

#### Combine the data from jan 2021 to dec 2022 into one data frame.

```{r}
bike_data <-rbind(                     
  read.csv("202112-divvy-tripdata.csv"),
  read.csv("202201-divvy-tripdata.csv"),
  read.csv("202202-divvy-tripdata.csv"),
  read.csv("202203-divvy-tripdata.csv"),
  read.csv("202204-divvy-tripdata.csv"),
  read.csv("202205-divvy-tripdata.csv"),
  read.csv("202206-divvy-tripdata.csv"),
  read.csv("202207-divvy-tripdata.csv"),
  read.csv("202208-divvy-tripdata.csv"),
  read.csv("202209-divvy-publictripdata.csv"),
  read.csv("202210-divvy-tripdata.csv"),
  read.csv("202211-divvy-tripdata.csv"),
  read.csv("202212-divvy-tripdata.csv"))
```

#### Examine the data frame.

```{r}

  head(bike_data)
  dim(bike_data)
  colnames(bike_data)
  summary(bike_data)
  nrow(bike_data)
```

#### Check for NA values.

```{r}
sum(is.na(bike_data))
```

#### drop columns we don't need

```{r}
 bike_data <- bike_data %>% select(-c(start_lat, start_lng, end_lat, end_lng))
bike_data 
```

#### drop NA

```{r}
remove_missing(bike_data)
drop_na(bike_data)
```


#### creating ride_length column for duration of ride calculated from start and end time of rides

```{r}
bike_data$ride_length <- difftime(bike_data$ended_at,bike_data$started_at)
#converting ride_length second to minutes
bike_data$ride_length <- as.numeric(bike_data$ride_length)
bike_data$ride_length <- as.numeric(bike_data$ride_length/60)
bike_data
```

#### add column "day_of_week,", and calculate the day of the week that each ride started

```{r}
bike_data <- bike_data%>% mutate(day_of_week = weekdays(as.Date(bike_data$started_at)))
bike_data
```

#### optional

```{r}
bike_data$date <- as.Date(bike_data$started_at)  #default format is yyyy-mm-dd
bike_data$month <-  format(as.Date(bike_data$date), "%m")
bike_data$day <- format(as.Date(bike_data$date), "%d")
bike_data$year <- format(as.Date(bike_data$date), "%Y")
```

#### remove data error

```{r}
bike_data <- bike_data[bike_data$ride_length>0,]

sum(bike_data$ride_length > 86400)  
```

#### analyze

```{r}
max(bike_data$ride_length) #max ride length
```

```{r}
max(bike_data$ride_length) #min ride length
```

#### Compare members and casual users

```{r}
aggregate(bike_data$ride_length ~ bike_data$member_casual, FUN = mean)

```

```{r}
aggregate(bike_data$ride_length ~ bike_data$member_casual, FUN = median)
```

```{r}
aggregate(bike_data$ride_length ~ bike_data$member_casual, FUN = max)
```

```{r}
aggregate(bike_data$ride_length ~ bike_data$member_casual, FUN = min)
```

#### Average ride time by each day for members vs casual users

```{r}
aggregate(bike_data$ride_length ~ bike_data$member_casual + bike_data$day_of_week, FUN = mean)
```

#### analyze data by type and weekdays

```{r}
bike_data %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  
  group_by(member_casual, weekday) %>%  
  summarise(number_of_rides = n()							
            ,average_duration = mean(ride_length)) %>% 		
  arrange(member_casual, weekday)
```

#### visualize the number of ride by rider type on weekday

```{r}
bike_data %>%
  mutate(weekday = wday(started_at, label=TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(), average_duration=mean
  (ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x=weekday, y=number_of_rides, fill=member_casual))+
  geom_col(position="dodge")
```

#### visualize the average duration by rider type on weekday

```{r}
bike_data %>%
  mutate(weekday = wday(started_at, label=TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(), average_duration=mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x=weekday, y=average_duration , fill=member_casual))+
  geom_col(position="dodge")
```

#### riding choice during day of the week

```{r paged.print=TRUE}
ggplot(data = bike_data) +
  geom_bar(mapping = aes(x = factor(day_of_week), fill = rideable_type))+
  facet_wrap(~member_casual) +
  labs(title='riding choice during day of the week', x= 'day of the week' )+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.2, hjust=0.5))
```
